\cursedforest\ - A Random Forest Implementation for ``Big'' and ``Wide'' Data \\ Supplementary Information

### 2016-11-27 doc/publications/plos16/plos16.tex
changes to plos16.tex

Random forests are a variance reduction technique. They take unpruned tree models, which have a low bias but a high
variance, and by combining them reduce the variance. As the prediction error is the sum of the variance and the bias
squared, this strategy should lead to a low prediction error.

The average prediction error for an individual tree $h(X; B)$ is

\cite{Friedman.1997}

\begin{equation}
PE_t = E_B E_{X,Y} (Y-h(X; B))^2.
\end{equation}
Assume that, for all $B$, the tree is unbiased, i.e., $EY= E_X h(X; B)$. Then
\begin{equation}
PE_f \leq \rho PE_t
\end{equation}
where $\rho$ is the weighted correlation between residuals $(Y-h(X;B))$ and $(Y-h(X;B^\prime))$ for independent $B,B^\prime$.  
So what is required for accurate random forest regression is (i) low correlation between residuals of differing tree in
the forest, and (ii) low prediction error for the individual trees \cite{Segal.2004}. In addition to growing the trees
on bootstrapped samples, each node in the tree is split on a randomly selected subset of the variables.

@@

Biau {\it et al.}~\cite{Biau.2012} provide a proof that a random forest model will converge at a rate that depends on
the cardinality of the set $S$ of ``strong predictors'' rather than on the number of variables $p$. That is, given a
function $y=f(S)$ depending on a set of variables $S$, a random forest will converge to the true function $y$ with a
rate that depends on the size of $S$ and not the number of noise variables in the data set. If the size of $S$ is small
compared to the total number of variables, then the rate of convergence will be quicker. It is not clear what the size
of the set $S$ will be in a GWAs application. There are cases of small groups of SNPs with large effects and also of
conditions caused by the interaction of large groups of SNPs. 
\marginpar{check this. Denis}

\cite{Genuer.et.al.2010} and \cite{Diaz.and.Alvarez.2006} describe iterative schemes for doing variable selection using
the importance measure. See \cite{Chen.and.Ishwaran.2012} for a review of the area with particular reference to genomic
data.
\marginpar{ \cite{Chen.and.Ishwaran.2012} cite many papers on extensions of random forests to handle 
GWAS data. What do we need to do in this area? What about linkage disequilibrium? Denis}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% This section talks about the datasets
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Datasets}
\subsubsection{1000 Genomes Project}
We obtained the 1000 Genomes Project data as VCF files from their FTP site.  Each VCF file contains the variants for
every individual across one chromosome.  We use the phase 3 dataset which contains 2,504 individuals with approximately
81 million variant positions (i.e. a matrix of $2,504 \times 81$ million).  For compatibility with the majority of the
algorithms we demonstrate, we convert the original variant representations to double representations. 
\marginpar{are they stored a double or integer?}
 For example, we
store the absence of a variant (0\textbar0) as 0, a heterozygous variant (1\textbar0 or 0\textbar1)as 1,  and a
homozygous variant (1\textbar1) as 2.  This is not the only possible encoding.  \cite{Goldstein.et.al.2011} note that of
the encoding listed in table \ref{table:encodings}, the additive encoding will present the most difficulty for a random
forest model.

>>>>

Random forests are a variance reduction technique. They take unpruned tree models, which have a low bias but a high
variance, and by combining them reduce the variance.  For regression problems, as the prediction error is the sum of the
variance and the bias squared, this strategy should lead to a low prediction error.

For classifiction problems there is no obvious decomposition of classification error into an equivalent bias plus
variance form. A number of different additive decompositions of classification error have been considered
in an effort to understand the success of bagging. They all less transparent than the regression
case (see \cite{Friedman.1997} and citations therein).

In the case of a regression tree, the average prediction error for an individual tree $h(X; B)$ is
\begin{equation}
PE_t = E_B E_{X,Y} (Y-h(X; B))^2.
\end{equation}
Assume that, for all $B$, the tree is unbiased, i.e., $EY= E_X h(X; B)$. Then
\begin{equation}
PE_f \leq \rho PE_t
\end{equation}
where $\rho$ is the weighted correlation between residuals $(Y-h(X;B))$ and $(Y-h(X;B^\prime))$ for independent $B,B^\prime$.  
So what is required for accurate random forest regression is (i) low correlation between residuals of differing tree in
the forest, and (ii) low prediction error for the individual trees \cite{Segal.2004}. In addition to growing the trees
on bootstrapped samples, each node in the tree is split on a randomly selected subset of the variables.

@@

Biau {\it et al.}~\cite{Biau.2012} provide a proof that a random forest model will converge at a rate that depends on
the cardinality of the set $S$ of ``strong predictors'' rather than on the number of variables $p$. That is, given a
function $y=f(S)$ depending on a set of variables $S$, a random forest will converge to the true function $y$ with a
rate that depends on the size of $S$ and not the number of noise variables in the data set. If the size of $S$ is small
compared to the total number of variables, then the rate of convergence will be quicker. It is not clear what the size
of the set $S$ will be in a GWAs application. There are cases of small groups of SNPs with large effects and also of
conditions caused by the interaction of large groups of SNPs.

\cite{Genuer.et.al.2010} and \cite{Diaz.and.Alvarez.2006} describe iterative schemes for doing variable selection using
the importance measure. See \cite{Chen.and.Ishwaran.2012} for a review of the area with particular reference to genomic
data.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% This section talks about the datasets
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Datasets}
\subsubsection{1000 Genomes Project}
We obtained the 1000 Genomes Project data as VCF files from their FTP site.  Each VCF file contains the variants for
every individual across one chromosome.  We use the phase 3 dataset which contains 2,504 individuals with approximately
81 million variant positions (i.e. a matrix of $2,504 \times 81$ million).  For compatibility with the majority of the
algorithms we demonstrate, we convert the original variant representations to double representations.  For example, we
store the absence of a variant (0\textbar0) as 0, a heterozygous variant (1\textbar0 or 0\textbar1)as 1, and a
homozygous variant (1\textbar1) as 2.  This is not the only possible encoding.  \cite{Goldstein.et.al.2011} note that of
the encoding listed in table \ref{table:encodings}, the additive encoding will present the most difficulty for a random
forest model.
